Continuation Passing Style for Effect Handlers

### 2020-02-02 docs/article.tex
Correct typos in the article

    The Freak implementation is available \href{https://github.com/Tomatosoup97/freak}{\underline{here}},
    written purely in Haskell. While the paper provided a good overview of the
    language and the translation, the lower-level details were omitted. That
    being said, two inherently different takes at the implementations were made.
    The first one is based on curried translation and A. Appel \cite{appel-continuations}
    book, and the second one directly on the uncurried translation to target
    calculus with continuations represented as explicit stacks from the paper.
    We start by presenting core data structrues, and afterwards move to actual
    translation details.

@@

    The first take was heavily inspired by A. Appel's Compiling with Continuations
    \cite{appel-continuations}, which provides a translation for a simplified ML
    calculus. The calculus was extended and translation adopted to handle algebraic
    effects and their handlers. The translation is based on the curried first-order
    translation. That being said, the source code diverged a lot from the paper on
    which it was based, leading to a different transformation for which
    the correctness and cohesion with operational semantics should be proved
    separately. Indeed, while the interpreter worked well on the use cases
    defined in tests, the evaluation had part which was not tail-recursive. What's
    more, nested handlers were not supported, and the implementation was found to be
    more tricky than it should, as it was not obvious on how to adopt the technique
    proposed in the paper. \\

    In terms of improving the performance of the evaluation, uncurried higher-order
    translation should be adopted, so that administrative redexes are contracted
    and proper tail-recursion is obtained. The core data structure, into which the
    source program is transformed, is defined as follows:

    \begin{verbatim}
data ContComp
    = CPSApp CValue [CValue]
    | CPSResume CValue ContComp
    | CPSFix Var [Var] ContComp ContComp
    | CPSBinOp BinaryOp CValue CValue Var ContComp
    | CPSValue CValue
    | CPSLet Var CValue ContComp
    | CPSSplit Label Var Var CValue ContComp
    | CPSCase CValue Label Var ContComp Var ContComp
    | CPSIf CValue ContComp ContComp
    | CPSAbsurd CValue
    \end{verbatim}
    Each term at the end has a coinductive reference to itself, which represents
    the rest of the computation that needs to be done. For more clarification,
    one may take a look that the book cited above. The source code for curried
    translation and evaulation can be found respetively in \verb!CPSAppel.hs! and
    \verb!EvalCPS.hs!.

@@

    This results in an implementation that finally supports nested handlers,
    as can be seen by evaluating \verb!programs/complexNestedHandlers.fk! program.
    Unfortunatelly, following closely translation from the paper resulted in
    behaviour in which invoked resumption forgets its pure continuation. This means,
    that the following code, evaluating correctly on the Appel-based translation,
    returns 0 rather than 1:

    \begin{verbatim}
handle do Drop () with { Drop p r -> let t <- r 0 in return 1}
    \end{verbatim}
    Nevertheless, working out this issue seems less demanding than coping with
    discrepancies created in the first translation. The source code for uncurried
    translation and evaulation can be found respetively in \verb!CPSLists.hs! and
    \verb!EvalTarget.hs!.

@@

    These implications are straightforward when looking from a categorical
    standpoint, where effects are viewed as free models of algebraic theories
    \cite{adequacy}, and handlers are homomorhpisms preserving the model structure
    \cite{handlers}. Nevertheless, the results are very exciting for programming
    use cases. The current Freak implementation does not support I/O.

@@

    The Freak language is experimental and a lot of possible enhancements
    could be adopted. Here we provide a list of the proposed directions in which
    further work could be done.

>>>>

    The Freak implementation is available \href{https://github.com/Tomatosoup97/freak}{\underline{here}},
    written purely in Haskell. While the paper provided a good overview of the
    language and the translation, the lower-level details were omitted. That
    being said, two inherently different takes at the implementations were made.
    The first one is based on curried translation and A. Appel \cite{appel-continuations}
    book, and the second one directly on the uncurried translation to target
    calculus with continuations represented as explicit stacks from the paper.
    We start by presenting core data structures, and afterwards move to actual
    translation details.

@@

    The first take was heavily inspired by A. Appel's Compiling with Continuations
    \cite{appel-continuations}, which provides a translation for a simplified ML
    calculus. The calculus was extended and translation adapted to handle algebraic
    effects and their handlers. The translation is based on the curried first-order
    translation. That being said, the source code diverged a lot from the paper on
    which it was based, leading to a different transformation for which
    the correctness and cohesion with operational semantics should be proved
    separately. Indeed, while the interpreter worked well on the use cases
    defined in tests, the evaluation had a part which was not tail-recursive. What's
    more, nested handlers were not supported, and the implementation was found to be
    trickier than it should, as it was not obvious on how to adopt the technique
    proposed in the paper. \\

    In terms of improving the performance of the evaluation, uncurried higher-order
    translation should be adapted, so that administrative redexes are contracted
    and proper tail-recursion is obtained. The core data structure, into which the
    source program is transformed, is defined as follows:

    \begin{verbatim}
data ContComp
    = CPSApp CValue [CValue]
    | CPSResume CValue ContComp
    | CPSFix Var [Var] ContComp ContComp
    | CPSBinOp BinaryOp CValue CValue Var ContComp
    | CPSValue CValue
    | CPSLet Var CValue ContComp
    | CPSSplit Label Var Var CValue ContComp
    | CPSCase CValue Label Var ContComp Var ContComp
    | CPSIf CValue ContComp ContComp
    | CPSAbsurd CValue
    \end{verbatim}
    Each term at the end has a coinductive reference to itself, which represents
    the rest of the computation that needs to be done. For more clarification,
    one may take a look that the book cited above. The source code for curried
    translation and evaluation can be found respectively in \verb!CPSAppel.hs! and
    \verb!EvalCPS.hs!.

@@

    This results in an implementation that finally supports nested handlers,
    as can be seen by evaluating \verb!programs/complexNestedHandlers.fk! program.
    Unfortunately, following closely translation from the paper resulted in
    behaviour in which invoked resumption forgets its pure continuation. This means,
    that the following code, evaluating correctly on the Appel-based translation,
    returns 0 rather than 1:

    \begin{verbatim}
handle do Drop () with { Drop p r -> let t <- r 0 in return 1}
    \end{verbatim}
    Nevertheless, working out this issue appears as less demanding than coping with
    discrepancies created in the first translation. The source code for uncurried
    translation and evaluation can be found respectively in \verb!CPSLists.hs! and
    \verb!EvalTarget.hs!.

@@

    These implications are straightforward when looking from a categorical
    standpoint, where effects are viewed as free models of algebraic theories
    \cite{adequacy}, and handlers are homomorphisms preserving the model structure
    \cite{handlers}. Nevertheless, the results are very exciting for programming
    use cases. The current Freak implementation does not support I/O.

@@

    The Freak language is experimental and a lot of possible enhancements
    could be adapted. Here we provide a list of the proposed directions in which
    further work could be done.
